---
title: "08-Setting_up_for_Success_with_golem"
author: "Ryan L. Metcalf"
date: "10/6/2021"
output: ioslides_presentation
---
# Setting Up for Success with {golem}

- **Learning Objectives**
 - Initialize Golem framework
 - Populate the framework with details
 - Review the Folder Structure
 - Comprehend how `here` works
 
## Setting up for Success with {golem}
- Developer Workflow:

1. PM sets up golem project
 - first steps filled
 - general structure is set
 - project is registered to version control
2. Start to prototyping the UI
 - work on CSS and JS elements
 - work on back-end functionalities
3. Work on integration of everything inside reacitve context

## Create a {golem}
1. Install libraries (if required)
2. File -> New Project -> Select "Package for shiny app Using golem"

Alternative, see code snippet below. Replace `path/to/package` with your known namespace.

```{r 08-step-by-step-prototype-1, eval = FALSE}
library(here)
library(golem)

# Creating a golem project from the command line
golem::create_golem(path = "~/Documents/GitHub/bookclub-epgs/golex")
```

You can view your new project directory structure using:

```{r 08-step-by-step-prototype-3, comment="", eval = FALSE}
# This is what a default {golem} project looks like
# Listing the files from the `golex` project using {fs}
fs::dir_tree("golex")
```

> NOTE: If you get and output with `fs::dir_tree("golex")` your package is working properly

## Setting things up with `dev/01_start.R`
`dev/01_start.R` is the first file that you need to populate and run.

### Fill the `DESCRIPTION` and set options
1. Within the `dev/01_start.R` file, you need to populate some pertinent information related to your project.

>NOTE: This is an automated system. Only edit and run this file first as the remaining steps will be completed in sequence for proper packaging process.

```{r 08-step-by-step-prototype-4, eval = FALSE}
golem::fill_desc(
  # The Name of the package containing the App 
  pkg_name = "ipsumapp", 
  # The Title of the package containing the App 
  pkg_title = "PKG_TITLE", 
  # The Description of the package containing the App 
  pkg_description = "PKG_DESC.", 
  # Your First Name
  author_first_name = "AUTHOR_FIRST", 
  # Your Last Name
  author_last_name = "AUTHOR_LAST", 
  # Your Email
  author_email = "AUTHOR@MAIL.COM", 
  # The URL of the GitHub Repo (optional) 
  repo_url = NULL 
)   
```

2. Run `golem::set_golem_options()` function.

>NOTE: This adds your information to the `golem-config.yml` file and sets the `here` package root sentinel. This is important as `here` manages the `/root` of your project. If if you change your working directory in the project, you will still be able to create modules and CSS files in their correct folders.

### Set common files
If you'd like to set your: Liscense, README, Code of Conduct, lifecycle badge, and NEWS,you can do so here.

```{r 08-step-by-step-prototype-5, eval = FALSE}
# You can set another license here
usethis::use_mit_license( name = "Golem User" )  
# Add a README, Code of Conduct, lifecycle badge and NEWS.md 
# file to your application
usethis::use_readme_rmd( open = FALSE )
usethis::use_code_of_conduct()
usethis::use_lifecycle_badge( "Experimental" )
usethis::use_news_md( open = FALSE )
```

Set up your version control account
```{r 08-step-by-step-prototype-6, eval = FALSE}
usethis::use_git()
```

### Use recommended elements
`golem::use_recommended_tests()` and `golem::use_recommended_deps()` sets a default testing infrastructure and adds dependencies to the application.

### Add utility functions
The next two code snippets (steps) will add various functions which can be used along the process of building your app.

```{r 08-step-by-step-prototype-7, eval = FALSE}
# These files will create R/golem_utils_ui.R 
# and R/golem_utils_server.R
golem::use_utils_ui()
golem::use_utils_server()
```

For example, using these utilities `list_to_li()` is a function to turn R lists into an HTML list. Another example is `with_red_star()` which puts a red astrix next to a mandatory input.

### Changing the favicon
You have an option of changing icons associated with your app. By default is the golem hex. Should you provide your own favicon library, you can use this code snippet:

```{r 08-step-by-step-prototype-8, eval = FALSE}
golem::use_favicon( path = "path/to/favicon")
```

Once complete, you can move to the second step, `dev/02_dev.R`

## Setting infrastructure for prototyping

### Add modules in `dev/02_dev.R`
The `golem::add_module()` function creates a module in the R folder. Each module created will follow the convention `mod_` and `mod_*_ui` and `mod*server` to the UI and server functions.

```{r 08-step-by-step-prototype-9, eval = FALSE}
# Creating a module skeleton
golem::add_module(name = "my_first_module") 
```

Here, you will paste (at the end of the file) that has to be copied and pasted inside your UI and server functions.

This is the core of your application and where the development focus begins.

### Add CSS and JS files
If you have particular "look and feel" of your app in the form