# Build Yourself a Safety Net

## Learning Objectives

## Tools and Resources

- {testthat}
- shiny::testServer
- puppeteer
- Puppeteer Recorder
- {crrri}
- {chromote}
- {crrry}
- gremlin.js
- {shinyloadtest} & shinycannon
- {shinytest}
- {dockerstats}
- {renv}
- Docker

## Testing Your App

What to test:

- business logic

- user interface

- reactive connections

- application load

### Testing the business logic

Good practice:

- business logic (non-reactive) is separated from interactive logic

- application is inside a package

Standard package development tools are available to you

- testthat

- devtools

```{r 11-01}
library(testthat)

meaning_of_life <- function() 42

describe("The meaning of life", {
  it("is always 42", {
    expect_equal(meaning_of_life(), 42)
  })  
})
```

### shiny::testServer

For testing reactive updates to server-side values

```{r 11-02}
library(shiny)

# Given the following module
computation_module_server <- function(id, multiplier = 10) {
  moduleServer(id, function(input, output, session) {
    ns <- NS(id)
    r <- reactiveValues(
      value = NULL
    )
    observeEvent(input$selector, {
      r$value <- input$selector * multiplier
    })
  })
}

describe("'value'", {
  it("updates to 'multiplier' * 'selector'", {
    # with default multiplier
    testServer(computation_module_server, {
      session$setInputs(selector = 1)
      expect_equal(r$value, 10)

      session$setInputs(selector = 2)
      expect_equal(r$value, 20)
    })
    # setting a non-default multiplier
    testServer(computation_module_server, args = list(multiplier = 15), {
      session$setInputs(selector = 3)
      expect_equal(r$value, 45)
    })
  })
})
```

### Testing the interactive logic

Several options available for testing UI and interactivity

#### Puppeteer

Mimics a session on the app

- [Puppeteer](https://developers.google.com/web/tools/puppeteer)
- NodeJS module
- Google Chrome headless session
- `npm install puppeteer`

Chrome extension

- [Headless recorder](https://chrome.google.com/webstore/detail/headless-recorder/djeegiggegleadkkbgopoonhjimgehda)
- [Website](https://www.checklyhq.com/docs/headless-recorder/)
- Creates scripts for puppeteer and playwright
- This records button clicks and text input
- Note: you have to hit TAB after text input or it won't be recorded

[Interactive]

- Load [Hexmake website](https://connect.thinkr.fr/hexmake/)
- Load [Headless Recorder "Basic Usage"](https://www.checklyhq.com/docs/headless-recorder/basic-usage/)
- Click "Headless Recorder" Icon in Chrome extensions
- Click the BIG RED BUTTON
- Click "Manage Name" on hexmake page
- Change Name to "dupree" then hit TAB
- Stop the recording
- Copy the code to the clipboard

```javascript
const puppeteer = require('puppeteer');
const browser = await puppeteer.launch()
const page = await browser.newPage()
await page.goto('https://connect.thinkr.fr/hexmake/')

await page.setViewport({ width: 898, height: 926 })

await page.waitForSelector('.row > .col > .rounded > details:nth-child(3) > summary')
await page.click('.row > .col > .rounded > details:nth-child(3) > summary')

await page.waitForSelector('#main_ui_1-left_ui_1-pkg_name_ui_1-package')
await page.click('#main_ui_1-left_ui_1-pkg_name_ui_1-package')

await page.type('#main_ui_1-left_ui_1-pkg_name_ui_1-package', 'dupree')

await browser.close()
```

```bash
# rerun the script
node ./examples/ch11/my_puppeteer_script.js
/home/russ/github/bookclub-epgs/examples/ch11/my_puppeteer_script.js:2
const browser = await puppeteer.launch()
                ^^^^^

SyntaxError: await is only valid in async functions and the top level bodies of modules
    at Object.compileFunction (node:vm:353:18)
    at wrapSafe (node:internal/modules/cjs/loader:1039:15)
    at Module._compile (node:internal/modules/cjs/loader:1073:27)
    at Object.Module._extensions..js (node:internal/modules/cjs/loader:1138:10)
    at Module.load (node:internal/modules/cjs/loader:989:32)
    at Function.Module._load (node:internal/modules/cjs/loader:829:14)
    at Function.executeUserEntryPoint [as runMain] (node:internal/modules/run_main:76:12)
```

This didn't actually work.
Comparing to the book example, and some [other puppeteer examples](https://nitayneeman.com/posts/getting-to-know-puppeteer-using-practical-examples/), 
we see it should probably have looked like this:

```javascript
const puppeteer = require('puppeteer');
(async () => {
const browser = await puppeteer.launch()
const page = await browser.newPage()
await page.goto('https://connect.thinkr.fr/hexmake/')

await page.setViewport({ width: 898, height: 926 })

await page.waitForSelector('.row > .col > .rounded > details:nth-child(3) > summary')
await page.click('.row > .col > .rounded > details:nth-child(3) > summary')

await page.waitForSelector('#main_ui_1-left_ui_1-pkg_name_ui_1-package')
await page.click('#main_ui_1-left_ui_1-pkg_name_ui_1-package')

await page.type('#main_ui_1-left_ui_1-pkg_name_ui_1-package', 'dupree')

await browser.close()
})()
```

```bash
node ./examples/ch11/my_fixed_puppeteer_script.js
```

[Interactive]

- Try changing the script:
  - use a non-existing selector and see the script fails
  - use `puppeteer.lauch({ headless: false });`

#### {crrri} and {crrry}

That's fine, but do we really want to have a node.js env in all our shiny projects?

- [{crrri}](https://rlesur.github.io/crrri/)
- [{chromote}](https://github.com/rstudio/chromote)
- [{crrry}](https://github.com/ColinFay/crrry) - puppeteer like, tailored for {shiny}

Why the strange names?

- they use the "Chrome remote interface"

```r
# to install {crrry}
remotes::install_github("ColinFay/crrry")
```

```r
# Creating a new test instance
test <- crrry::CrrryOnPage$new(
  # Using the `find_chrome()` function to guess where the 
  # Chrome bin is on our machine
  chrome_bin = pagedown::find_chrome(),
  # Launching Chrome on a random available port on our machine
  # Note that you will need httpuv >= 1.5.2 if you want to use 
  # this function
  chrome_port = httpuv::randomPort(), 
  # Specifying the page we want to connect to
  url = "https://connect.thinkr.fr/hexmake/",
  # Do everything on the terminal, with no window open
  headless = TRUE
)
```

```r
# We'll wait for the application to be ready to accept inputs
test$wait_for_shiny_ready()
```

A simple test:

- we input a package name
- then check that the updated package name matches our input name

```r
new_pkg <- "dupree"
pkg_name_selector <- "main_ui_1-left_ui_1-pkg_name_ui_1-package"
js_get_pkg <- glue::glue("$('#{pkg_name_selector}').attr('value')")

test$shiny_set_input(pkg_name_selector, new_pkg)
test$wait_for_shiny_ready()

# This (seemingly trivial) test fails:
# - test$shiny_set_input doesn't change the 'visible' input
# - but, it triggers reactive changes on the server side
expect_equal(
  test$call_js(js_get_pkg)[["result"]][["value"]],
  expected = new_pkg
)
```

That test failed (`test$set_shiny_input` sets the inputs that are passed to the server).
We can use javascript / jquery syntax to set the UI-attached values.

```r
new_pkg <- "dupree"
pkg_name_selector <- "main_ui_1-left_ui_1-pkg_name_ui_1-package"
js_get_pkg <- glue::glue("$('#{pkg_name_selector}').attr('value')")
js_set_pkg <- glue::glue("$('#{pkg_name_selector}').attr('value', '{new_pkg}')")

test$call_js(js_set_pkg)
test$wait_for_shiny_ready()

expect_equal(
  test$call_js(js_get_pkg)[["result"]][["value"]],
  expected = new_pkg
)
```

```r
test$stop()
```

Available methods:

- call_js()
- shiny_set_input()
- wait_for()
- click_on_id

#### gremlin

#### {shinytest}

Similar to {puppeteer}-based tests

Uses snapshots (browser images) to compare before/after source code changes

[Interactive]

- open hexmake repo

## Further sections

Did not present on:

- Testing the app load

- A reproducible environment and {renv}

-  Docker

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/p5IXw-AlBK4")`

<details>
  <summary> Meeting chat log </summary>
```
00:35:42	Federica Gazzelloni:	puppeteer : https://developers.google.com/web/tools/puppeteer
00:36:17	Federica Gazzelloni:	testthat: https://testthat.r-lib.org/
00:39:34	Federica Gazzelloni:	hexmake: https://connect.thinkr.fr/hexmake/
00:40:21	Federica Gazzelloni:	headless-recorder: https://chrome.google.com/webstore/detail/headless-recorder/djeegiggegleadkkbgopoonhjimgehda
01:07:01	Federica Gazzelloni:	shinytest: https://rstudio.github.io/shinytest/
```
</details>
