# Optimizing {shiny} code

**Learning Objectives**

- Major routes for optimising R/shiny code

- Cacheing

- Asynchronous code

## Resources

- Efficient R programming [https://csgillespie.github.io/efficientR/]()

- Advanced R [https://adv-r.hadley.nz/perf-improve.html]()

- Mastering Shiny [https://mastering-shiny.org/performance.html]()

- RStudio caching blogposts:
  - [https://shiny.rstudio.com/articles/plot-caching.html]() - the old way
  - [https://shiny.rstudio.com/articles/caching.html]() - the new way

Packages

- {R.cache} [https://github.com/HenrikBengtsson/R.cache]()

- {memoise} [https://github.com/r-lib/memoise]()

- {future} [https://github.com/HenrikBengtsson/future]()

- {promises} [https://rstudio.github.io/promises]()

## Caching Elements

### What is Caching

- Store resource-intensive results

- So they can be reused without recomputation

- Downside: you can't cache 'em all

Only cache if the thing:

- would be identical each time it is made (for a given input)

Good things to cache:

- Plots

- Database queries

### Native caching in R

{R.cache} and {memoise} work in a similar way:

- Wrap a function with a higher-order function (`new_func <- memoise::memoise(initial_func)`)

- Call the new function with some arguments (`new_func(x = 1, y = 2)`)
  - it does the expensive computation
  - then stores the result

- Call the new function again
  - if you've used the args before, it will return immediately with the stored result

- The arguments are converted to a look-up key

```{r memoise_your_first_fn, eval = FALSE}
library(memoise)

# tic/toc are used for timing
library(tictoc)

sleep_and_return_time <- function(seconds = 1) {
  Sys.sleep(seconds)
  return(Sys.time())
}

msleep_and_return_time <- memoise(sleep_and_return_time)

# first call is sloooooooow
tic()
msleep_and_return_time(10)
toc()

# second call is fast
tic()
msleep_and_return_time(10)
toc()
```

But .... why is `sleep_and_return_time` a bad example for a cacheable function?

The book has a more reproducible example using database queries (there's a db-in-shiny example
below, so we don't cover this example)

{memoise} stores the cache on disk, you can set where on the file-system it is stored

### Caching in {shiny}

An app:

```r
library(shiny)

ui <- function(){
  tagList(
    # The user can select one of the cut from ggplot2::diamonds,
    # {shiny} will then query the SQL database to retrieve the
    # first rows of the result
    selectInput("cut", "cut", unique(ggplot2::diamonds$cut)),
    tableOutput("tbl")
  )
}

srv_builder <- function(con) {
  server <- function(
    input,
    output,
    session
  ){
    # Rendering the table of the SQL call
    output$tbl <- renderTable({
      # Using a memoised function allows to prevent from
      # calling the SQL database every time the user inputs
      # a change
      memoised_fct_sql(input$cut, con)
    })
  }

  server
}

# We create an in-memory database using SQLite
con <- DBI::dbConnect(
  RSQLite::SQLite(),
  dbname = ":memory:"
)

# Writing a large dataset to the db
DBI::dbWriteTable(
  con,
  "diams",
  # This table will have 539400 rows
  dplyr::bind_rows(
    purrr::rerun(10, ggplot2::diamonds)
  )
)

shinyApp(ui, server = srv_builder(con))
```

What haven't we defined here?

```r
library(memoise)

fct_sql <- function(cut, con){
  # NEVER EVER SPRINTF AN SQL CODE LIKE THAT
  # IT'S SENSITIVE TO SQL INJECTIONS, WE'RE
  # DOING IT FOR THE EXAMPLE
  cli::cat_rule("Calling the SQL db")
  results <- DBI::dbGetQuery(
    con, sprintf(
      "SELECT * FROM diams WHERE cut = '%s'",
      cut
    )
  )
  head(results)
}

# Using a local cache
cache_dir <- cache_filesystem("cache")
memoised_fct_sql <- memoise(fct_sql, cache = cache_dir)
```

App workflow:

- Start the app
- set input$cut to a new value
- time how long it takes to render the new table
- set input$cut to an unused value
- note that it takes the same length of time
- set input$cut to the first selected value
- note that it is way faster than the original call

{shiny} already has caching functionality:

- `shiny::bindCache()`
- [older] `shiny::renderCachedPlot()`

Note from the RStudio shiny [blog](https://shiny.rstudio.com/articles/plot-caching.html):

"""As of Shiny 1.6.0, we recommend using renderPlot() with bindCache() instead"""

Example shiny app

- taken from the book
- but rewritten to use `renderPlot() %>% bindCache()` rather than `renderCachedPlot()`

```{r shiny_cacheing, eval = FALSE}
library(shiny)

ui <- function() {
  tagList(
    # We select a data.frame to plot
    selectInput(
      "tbl",
      "Table",
      c("iris", "mtcars", "airquality")
    ),
    # This plotOutput will be cached
    plotOutput("plot")
  )
}

server <- function(
  input,
  output,
  session
) {
  # The cache mechanism is made available by 'bindCache'
  output$plot <- renderPlot({
    # Plotting the selected data.frame
    plot(get(input$tbl))
  }) %>%
    bindCache(
    # List here all the reactive expression that will
    # be used as cache key when running the app,
    # you will see that the first time you plot one
    # graph, it takes a couple of seconds,
    # but the second time, it's almost
    # instantaneous
    input$tbl
  )
}

shinyApp(ui, server)
```

Notes:

- you can also cache on remote storage (e.g., S3)
- you only have limited space - old cached values will be dropped

## Asynchronous in {shiny}

R is single threaded but the world isn't

- B and C depend on A
- D depends on B and C

Synchronous code:

- either A -> B -> C -> D
- or A -> C -> B -> D

If B runs on another computer and takes a long time

```
Synchronous version:

A -> start-B             retrieve-B -> C -> D
            \           /     (My-server)
             \         /      ---------------
              - run-B -       (Remote-server)
```

```
Asynchronous version:

A -> start-B -> C -> retrieve-B -> D
       \              /    (My-server)
        \            /     ---------------
         - run B ----      (Remote-server)
```

Packages:

- {future}: send computation elsewhere
- {promises}: objects for handling async computation

### Asynchronously ease cross-session issues

- One R/shiny session
- Multiple users
- A few, specific, long-running operations

"Promise" = A stand-in for the eventual result of an operation
  - [https://rstudio.github.io/promises/articles/intro.html]()
  - `%...>%`
    - Promise-compatible pipe
    - What happens on success
  - `%...!%`
    - What happens on failure

```r
library(shiny)
library(future)
library(promises)
plan(multisession)

ui <- function(){
  tagList(
    # This will receive the output of the future
    verbatimTextOutput("rnorm")
  )
}

server <- function(
  input,
  output,
  session
){
  output$rnorm <- renderPrint({
    # Sending the rnorm to be run in another session
    future({
      Sys.sleep(20)
      return(rnorm(5))
    }) %...>%
      print(.) %...!%
      stop(.)
  })
}

s <- shinyApp(ui, server)
```

```r
runApp(s, launch.browser = FALSE)
```

Plan:

- run the above
- launch the app in two separate browser windows (X and Y)
- wait 20 seconds
- If one session blocked the other:
  - X results would show at ~ 20secs
  - Y results at ~ 40secs
- If there is no cross-session blocking
  - Both X and Y results would show at ~ 20secs

### Asynchronously ease within-session issues

(TODO)

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary> Meeting chat log </summary>

```
CHAT LOG HERE
```
</details>
