# Using JavaScript

**Learning Objectives**

 - Interact with `{shiny}` apps using JavaScript `r emo::ji("rocket")` blasting your work into the ether
 - Manipulate UI and Server linkages with JavaScript
 - Learn the various methods to apply JavaScript to your `{shiny}` app
 - Understand how to reference elements of a HTML Document using JavaScript
 - Comprehend the differences between JavaScript and jQuery

## Introduction
This is a **HUGE** chapter!!! But then, again, JavaScript is a **HUGE** subject! JavaScript is maintained by the [European Computer Manufacturers Association (ECMA)](https://www.ecma-international.org/). JavaScript is often referred to as *ECMA Script*. 

Where do I use Javascript? Here are just a few examples:

 - Interactive Webpages (Radio Buttons, Text Entry, dynamic changing of page attributes)
 - eLearning (SCORM, xAPI)
 - Data Visualization (D3JS)
 - You are using it to Knit this Markdown as well as provide all sorts of interaction!
 
Javascript is one of three primary standards that make up the [World Wide Web Consortium (W3C)](https://www.w3.org/). Other standards include:

 - HyperText Markup Language (HTML)
 - Cascading Style Sheets (CSS)
 - Standard Vectors Graphics (SVG)
 - jQuery (a derivative of JavaScript. Considered more *lightweight*)
 - Asynchronous JavaScript And XML (AJAX)

This may all appear confusing, but the reality is, most R developers who deploy `{shiny}` apps may not even leave RStudio. Therefore, one could claim the primary reason for adding additional JavaScript is to add *hyper performance* to your web app!

>Note, at this point, I'm not calling it a `{shiny}` anymore, but a *web app*.

But what is going on under the hood?

- `{shiny}` binds events to UI elements, and these JavaScript events will communicate with R
- R talks to your browser through a web socket
   - Communication happens in both directions

## A quick introduction to JavaScript

- The fastest way to interact with JavaScript is from the Developer Tools of your favorite browser
- For example, enter the following in your Dev Tools Console: `var message = "Hello World"; alert(message);`

### Including JavaScript code in your app
There are three primary ways to include JavaScript code in your app:

- As an external file
- Inside a `<script>Your Code snippet goes here</script>` tag
- Inline, on a specific tag

The recommended practice is through an external file.

If you are developing with `{golem}`, there are two primary functions to include JavaScript. They are:

- `golem::add_js_file("name")`, this adds a standard JavaScript file. This method is not intended to communicate with R
- `golem::add_js_handler("name")`, this is intended to work with R
- `golem::add_js_binding("name")`, for advanced, customer interaction for your shiny app

### Understanding HTML, class, and id
Think of your Web Page as a tree where the top of the page is your root node.

- Referenced as the Document Object Model (DOM)
  - You can work on any of these HTML nodes with JavaScript

The following code snippets compare R code to its corresponding HTML output:

```{r, R Code, eval=FALSE}
library(shiny)
fluidPage(
  titlePanel("Hello Shiny"), 
   textInput("act", "Input")
)
```

And its corresponding HTML:

```{html, HTML code}
<div class="container-fluid">
  <h2>Hello Shiny</h2>
  <div class="form-group shiny-input-container">
    <label class="control-label" for="act">Input</label>
    <input id="act" type="text" class="form-control" value=""/>
  </div>
</div>
```

We can see we created a `div` tag with a Bootstrap container, an `h2` tag for text, and a text input field with and id and class.

>NOTE: `id` must be unique. This is how the websockets can communicate

Elements can have a class which can apply to multiple elements. This can be used in JavaScript, but it is also very useful for styling elements in CSS.

### Querying in Vanilla JavaScript
The term *Vanilla JavaScript* refers to plain JavaScript, meaning, not external plugins.

```{r, JavaScript examples, eval = FALSE}
// Given
<div id = "first" name="number" class = "widediv">Hey</div>

// Query with the ID
document.querySelector("#first")
document.getElementById("first")

// With the class
document.querySelectorAll(".widediv")
document.getElementsByClassName("widediv") 

// With the name attribute
document.getElementsByName("number") 

// Using the tag name
document.getElementsByTagName("div")
```

This method comes with a hitch! Not all Browsers are created equal! This brings another angle of development consideration to our software development process. With relation to proprietary frameworks and a host of different Browser vendors (Mozilla Firefox, Google Chrome, Apple Safari, etc...) not all JavaScript versions are recognized.

### About DOM Events
Within the Document Object Model, interactions are called *events*. JavaScript runtime is listening for these events. Better yet, this is were we can control the interaction of our users with our `{shiny}` apps. The following are examples of events:

- `click` / `dblclick`

- `focus`

- `keypress`, `keydown`, `keyup`

- `mousedown`, `mouseenter`, `mouseleave`, `mousemove`, `mouseout`, `mouseover`, `mouseup`

- `scroll`

For a full list of these events, please visit [this link](https://developer.mozilla.org/fr/docs/Web/Events).

Once you have this list in mind, you can select elements in the DOM, add an `addEventListener` to them, and define a callback function (which is executed when the event is triggered)

```{html, Key Press Example, eval = FALSE}
<input type="text" id = "firstinput">
<script> 
  document.getElementById("firstinput").addEventListener(
    "keypress", 
    function(){
      alert("Pressed!")
    }
  )
</script>
```

Another example is to send an alert when the Shiny Server is connected:

```{r, jQuery Alert Message, eval = FALSE}
$(document).on('shiny:connected', function(event) {
  alert('Connected to the server'); 
});
```

The `$()` is considered jQuery and is a subset, lighter weight, more widely recognized method of running JavaScript.

### About `jQuery` and `jQuery` selectors
jQuery is native to the `{shiny}` package. Examples of jQuery include:

- `$("#firstinput")` to refer to the element with the id `firstinput`

- `$(".widediv")` to refer to element(s) of class `widediv`

- `$("button:contains('this')")` to refer to the buttons with a text containing `'this'`

```{html, HTML Code Example, eval = FALSE}
<a href = "https://thinkr.fr" data-value = "panel2">ThinkR</a>
```

We can interact with these attributes using jQuery:

- `$("a[href = 'https://thinkr.fr']")` refers to link(s) with `href` being `https://thinkr.fr`

- `$('a[data-value="panel2"]')` refers to link(s) with `data-value` being `"panel2"`

>NOTE: jQuery has the advantage of simplifying selections and actions and is a cross-platform library

## Client-Side JavaScript
This whole topic is related to JavaScript, but is stated with relation to R and Shiny. These three are all intertwined with each other. If extend beyond R and `{shiny}`, you can expand the interaction of your web app. Examples include: Small JavaScript functions that will prevent you from writing complex algorithmic logic in your application server.

- `alert("message")` is a built-in alert-box mechanism. This is built into the user's browser and can replace a call to the server, like `shinyalert`.
 
- `var x = prompt("this", "that");` is a function that opens a built-in prompt window. We can recieve input and store it in a variable for future use.

### Common Patterns
Examples of Client-Side use of JavaScript:

- `alert("message")` is a built-in alert-box mechanism. This is built into the user's browser and can replace a call to the server, like `shinyalert`.
 
- `var x = prompt("this", "that");` is a function that opens a built-in prompt window. We can recieve input and store it in a variable for future use.

- `$('#id').css('color', 'green');`, or in vanilla JavaScript `document.getElementById("demo").style.color = "green";` changes the CSS attributes of the selected element(s). Here, we are switching to green on the `#id` element.

- `$("#id").text("this")`, or in vanilla JavaScript `document.getElementById("id").innerText = "this";` changes the text content to "this".

- `$("#id").remove();`, or in vanilla JavaScript `var elem = document.querySelector('#some-element'); elem.parentNode.removeChild(elem);` completely removes the element from the DOM. It can be used as a replacement for `shiny::removeUI()`, or as a conditional UI. 

> Note that this code doesn't remove the input values on the server side: the elements only disappear from the UI, but nothing is sent to the server side.

### Where to put them: Back to JavaScript Events
It is great we are discussing all levels of JavaScript implementation...but where to put them?

Here are some examples of adding JavaScript functions to DOM events:

-   `onclick`

The `onclick` attribute can be added straight inside the HTML tag when possible:

```{r, OneClick Example, eval = FALSE}
# Building a button using the native HTML tag 
# (i.e. not using the actionButton() function)
# This button only goal is to launch this JS code
# when it is clicked 
tags$button(
  "Show",
  onclick = "$('#plot').show()"
)
```

Or with `shiny::tagAppendAttributes()`:

```{r, Shiny Append Attributes Example, eval = FALSE}
# Using tagAppendAttributes() allows to add attributes to the 
# outputed UI element
plotOutput(
  "plot"
) %>% tagAppendAttributes(
  onclick = "alert('hello world')"
)
```

Here is, for example, a small `{shiny}` app that implements this behavior:

```{r, Shiny with Magrittr, eval = FALSE}
library(shiny)
library(magrittr)
ui <- function(){
  fluidPage(
    # We create a plotOutput, which will show an alert when 
    # it is clicked
    plotOutput(
      "plot"
    ) %>% tagAppendAttributes(
      onclick = "alert('iris plot!')"
    )
  )
}
server <- function(input, output, session){
  output$plot <- renderPlot({
    plot(iris)
  })
}
shinyApp(ui, server)
```

**Lets put this into Golem!!!**

Use the `add_js_file("name")` function in Golem.

- In `/inst/app/www/script.js`

```{js, Golem Javascript example, eval=FALSE}
function alertme(id){
  // Asking information
  var name = prompt("Who are you?");
  // Showing an alert
  alert("Hello " + name + "! You're seeing " + id);
}
```

- Then in R

```{r, Golem R Code Example, eval=FALSE, echo=TRUE}
plotOutput(
  "plot"
) %>% tagAppendAttributes(
  # Calling the function which has been defined in the 
  # external script
  onclick = "alertme('plot')"
)
```

Last example, you can also add jQuery calls too! Inside this `inst/app/www/script.js`, you can also attach a new behavior with `jQuery` to one or several elements.
For example, you can add this `alertme` / `onclick` behavior to all plots of the app:

```{js, jQuery example, eval = FALSE, echo = TRUE}
function alertme(id){
  var name = prompt("Who are you?");
  alert("Hello " + name + "! You're seeing " + id);
}
/* We're adding this so that the function is launched only
when the document is ready */
$(function(){ 
  // Selecting all `{shiny}` plots
  $(".shiny-plot-output").on("click", function(){
    /* Calling the alertme function with the id 
    of the clicked plot */
    alertme(this.id);
  });
});
```

We've unpacked a lot here...but there is more. Check out [JavaScript Events in `{shiny}`](https://shiny.rstudio.com/articles/js-events.html) for the full list of JavaScript events available in `{shiny}`.

## JavaScript <-> `{shiny}` Communication
Now that we have seen some client-side optimization, i.e. R does not do anything with these events when they happen (in fact R is not even aware they happened).

### From R to JavaScript
Calling JS from the server side (i.e. from R) is done by defining a series of `CustomMessageHandler` functions: these are functions with one argument that can then be called using the `session$sendCustomMessage()` method from the server side.Or if you are using `{golem}`, using the `invoke_js()` function.

Using the skeleton from `golem::add_js_handler("first_handler")`, we get this snippet:

```{r, Another jQuery snippet, eval=FALSE}
$( document ).ready(function() {
  Shiny.addCustomMessageHandler('fun', function(arg) {
  
  })
});
```

Then, you can call it from the server side:

```{r, Server Side Script Example, eval = FALSE}
session$sendCustomMessage("fun", list())
# OR
golem::invoke_js("fun", ...)
```

>NOTE: the `list()` argument will be coverted to JavaScript Object Notation (JSON) and read via JavaScript. If we have function `x` and variables `a` and `b`, then we can use the dot notation of `x.a` and `x.b` in our JavaScript.

For Example:

```{js, Dot Notation, eval = FALSE}
// We define a handler called "computed", that can be called 
// from the server side of the {shiny} application
Shiny.addCustomMessageHandler('computed', function(mess) {
  // The received value (in mess) is serialized in JSON, 
  // so we can  access the list element with object.name
  alert("Computed " + mess.what + " in " + mess.sec + " secs");
})
```

And within R:

```{r, Server side using Dot Notation, eval = FALSE}
observe({
  # Register the starting time
  deb <- Sys.time()
  # Mimic a long computation
  Sys.sleep(
    sample(1:5, 1)
  )
  # Calling the computed handler
  golem::invoke_js(
    "computed", 
    # We send a list, that will be turned into JSON
    list(
      what = "time", 
      sec = round(Sys.time() - deb)
    )
  )
})
```

### From JavaScript to R
>NOTE: My presentation ends here.....I need to finish the remaining three sections.

## About {shinyjs} JS functions
PUT SOME TEXT HERE

## One last thing: API calls
PUT SOME TEXT HERE

## Learn more about JavaScript
Here is a list of external resources to learn more about JavaScript:

### `{shiny}` and JavaScript

-   We have written an online, freely available book about `{shiny}` and JavaScript: [_JavaScript 4 `{shiny}` - Field Notes_](http://connect.thinkr.fr/js4shinyfieldnotes/).

-   [JavaScript for `{shiny}` Users](https://js4shiny.com/), companion website to the rstudio::conf(2020) workshop.

-   [Build custom input objects](https://shiny.rstudio.com/articles/building-inputs.html).

-   [Packaging JavaScript code for `{shiny}`](https://shiny.rstudio.com/articles/packaging-javascript.html).

-   [Communicating with `{shiny}` via JavaScript](https://shiny.rstudio.com/articles/communicating-with-js.html).

### JavaScript basics

-   [Mozilla JavaScript](https://developer.mozilla.org/en-US/docs/Web/JavaScript)
-   [w3schools JavaScript](https://www.w3schools.com/js/default.asp)
-   [Free Code Camp](https://www.freecodecamp.org/)
-   [JavaScript for Cats](http://jsforcats.com/)
-   [Learn JS](https://www.learn-js.org/)

### jQuery

-   [jQuery Learning Center](https://learn.jquery.com/)
-   [w3schools jQuery](https://www.w3schools.com/jquery/default.asp)

### Intermediate/advanced JavaScript

-   [Eloquent JavaScript](https://eloquentjavascript.net/)
-   [You Don't Know JS Yet](https://github.com/getify/You-Dont-Know-JS)