# Using JavaScript

**Learning Objectives**

 - Interact with `{shiny}` apps using JavaScript `r emo::ji("rocket")` blasting your work into the ether
 - Manipulate UI and Server linkages with JavaScript
 - Learn the various methods to apply JavaScript to your `{shiny}` app
 - Understand how to reference elements of a HTML Document using JavaScript
 - Comprehend the differences between JavaScript and jQuery

## Introduction
This is a **HUGE** chapter!!! But then, again, JavaScript is a **HUGE** subject! JavaScript is maintained by the [European Computer Manufacturers Association (ECMA)](https://www.ecma-international.org/). JavaScript is often referred to as *ECMA Script*. 

Where do I use Javascript? Here are just a few examples:

 - Interactive Webpages (Radio Buttons, Text Entry, dynamic changing of page attributes)
 - eLearning (SCORM, xAPI)
 - Data Visualization (D3JS)
 - You are using it to Knit this Markdown as well as provide all sorts of interaction!
 
Javascript is one of three primary standards that make up the [World Wide Web Consortium (W3C)](https://www.w3.org/). Other standards include:

 - HyperText Markup Language (HTML)
 - Cascading Style Sheets (CSS)
 - Standard Vectors Graphics (SVG)
 - jQuery (a derivative of JavaScript. Considered more *lightweight*)
 - Asynchronous JavaScript And XML (AJAX)

This may all appear confusing, but the reality is, most R developers who deploy `{shiny}` apps may not even leave RStudio. Therefore, one could claim the primary reason for adding additional JavaScript is to add *hyper performance* to your web app!

>Note, at this point, I'm not calling it a `{shiny}` anymore, but a *web app*.

But what is going on under the hood?

- `{shiny}` binds events to UI elements, and these JavaScript events will communicate with R
- R talks to your browser through a web socket
   - Communication happens in both directions

## A quick introduction to JavaScript

- The fastest way to interact with JavaScript is from the Developer Tools of your favorite browser
- For example, enter the following in your Dev Tools Console: `var message = "Hello World"; alert(message);`

### Including JavaScript code in your app
There are three primary ways to include JavaScript code in your app:

- As an external file
- Inside a `<script>Your Code snippet goes here</script>` tag
- Inline, on a specific tag

The recommended practice is through an external file.

If you are developing with `{golem}`, there are two primary functions to include JavaScript. They are:

- `golem::add_js_file("name")`, this adds a standard JavaScript file. This method is not intended to communicate with R
- `golem::add_js_handler("name")`, this is intended to work with R
- `golem::add_js_binding("name")`, for advanced, customer interaction for your shiny app

### Understanding HTML, class, and id
Think of your Web Page as a tree where the top of the page is your root node.

- Referenced as the Document Object Model (DOM)
  - You can work on any of these HTML nodes with JavaScript

The following code snippets compare R code to its corresponding HTML output:

```{r, R Code, eval=FALSE}
library(shiny)
fluidPage(
  titlePanel("Hello Shiny"), 
   textInput("act", "Input")
)
```

And its corresponding HTML:

```{html, HTML code}
<div class="container-fluid">
  <h2>Hello Shiny</h2>
  <div class="form-group shiny-input-container">
    <label class="control-label" for="act">Input</label>
    <input id="act" type="text" class="form-control" value=""/>
  </div>
</div>
```

We can see we created a `div` tag with a Bootstrap container, an `h2` tag for text, and a text input field with and id and class.

>NOTE: `id` must be unique. This is how the websockets can communicate

Elements can have a class which can apply to multiple elements. This can be used in JavaScript, but it is also very useful for styling elements in CSS.

### Querying in Vanilla JavaScript
The term *Vanilla JavaScript* refers to plain JavaScript, meaning, not external plugins.

```{r, JavaScript examples, eval = FALSE}
// Given
<div id = "first" name="number" class = "widediv">Hey</div>

// Query with the ID
document.querySelector("#first")
document.getElementById("first")

// With the class
document.querySelectorAll(".widediv")
document.getElementsByClassName("widediv") 

// With the name attribute
document.getElementsByName("number") 

// Using the tag name
document.getElementsByTagName("div")
```

This method comes with a hitch! Not all Browsers are created equal! This brings another angle of development consideration to our software development process. With relation to proprietary frameworks and a host of different Browser vendors (Mozilla Firefox, Google Chrome, Apple Safari, etc...) not all JavaScript versions are recognized.

### About DOM Events
Within the Document Object Model, interactions are called *events*. JavaScript runtime is listening for these events. Better yet, this is were we can control the interaction of our users with our `{shiny}` apps. The following are examples of events:

- `click` / `dblclick`

- `focus`

- `keypress`, `keydown`, `keyup`

- `mousedown`, `mouseenter`, `mouseleave`, `mousemove`, `mouseout`, `mouseover`, `mouseup`

- `scroll`

For a full list of these events, please visit [this link](https://developer.mozilla.org/fr/docs/Web/Events).

Once you have this list in mind, you can select elements in the DOM, add an `addEventListener` to them, and define a callback function (which is executed when the event is triggered)

```{html, Key Press Example, eval = FALSE}
<input type="text" id = "firstinput">
<script> 
  document.getElementById("firstinput").addEventListener(
    "keypress", 
    function(){
      alert("Pressed!")
    }
  )
</script>
```

Another example is to send an alert when the Shiny Server is connected:

```{r, jQuery Alert Message, eval = FALSE}
$(document).on('shiny:connected', function(event) {
  alert('Connected to the server'); 
});
```

The `$()` is considered jQuery and is a subset, lighter weight, more widely recognized method of running JavaScript.

### About `jQuery` and `jQuery` selectors
jQuery is native to the `{shiny}` package. Examples of jQuery include:

- `$("#firstinput")` to refer to the element with the id `firstinput`

- `$(".widediv")` to refer to element(s) of class `widediv`

- `$("button:contains('this')")` to refer to the buttons with a text containing `'this'`

```{html, HTML Code Example, eval = FALSE}
<a href = "https://thinkr.fr" data-value = "panel2">ThinkR</a>
```

We can interact with these attributes using jQuery:

- `$("a[href = 'https://thinkr.fr']")` refers to link(s) with `href` being `https://thinkr.fr`

- `$('a[data-value="panel2"]')` refers to link(s) with `data-value` being `"panel2"`

>NOTE: jQuery has the advantage of simplifying selections and actions and is a cross-platform library

## Client-Side JavaScript
This whole topic is related to JavaScript, but is stated with relation to R and Shiny. These three are all intertwined with each other. If extend beyond R and `{shiny}`, you can expand the interaction of your web app. Examples include: Small JavaScript functions that will prevent you from writing complex algorithmic logic in your application server.

- `alert("message")` is a built-in alert-box mechanism. This is built into the user's browser and can replace a call to the server, like `shinyalert`.
 
- `var x = prompt("this", "that");` is a function that opens a built-in prompt window. We can recieve input and store it in a variable for future use.

### Common Patterns
Examples of Client-Side use of JavaScript:

- `alert("message")` is a built-in alert-box mechanism. This is built into the user's browser and can replace a call to the server, like `shinyalert`.
 
- `var x = prompt("this", "that");` is a function that opens a built-in prompt window. We can recieve input and store it in a variable for future use.

- `$('#id').css('color', 'green');`, or in vanilla JavaScript `document.getElementById("demo").style.color = "green";` changes the CSS attributes of the selected element(s). Here, we are switching to green on the `#id` element.

- `$("#id").text("this")`, or in vanilla JavaScript `document.getElementById("id").innerText = "this";` changes the text content to "this".

- `$("#id").remove();`, or in vanilla JavaScript `var elem = document.querySelector('#some-element'); elem.parentNode.removeChild(elem);` completely removes the element from the DOM. It can be used as a replacement for `shiny::removeUI()`, or as a conditional UI. 

> Note that this code doesn't remove the input values on the server side: the elements only disappear from the UI, but nothing is sent to the server side.

### Where to put them: Back to JavaScript Events
It is great we are discussing all levels of JavaScript implementation...but where to put them?

Here are some examples of adding JavaScript functions to DOM events:

-   `onclick`

The `onclick` attribute can be added straight inside the HTML tag when possible:

```{r, OneClick Example, eval = FALSE}
# Building a button using the native HTML tag 
# (i.e. not using the actionButton() function)
# This button only goal is to launch this JS code
# when it is clicked 
tags$button(
  "Show",
  onclick = "$('#plot').show()"
)
```

Or with `shiny::tagAppendAttributes()`:

```{r, Shiny Append Attributes Example, eval = FALSE}
# Using tagAppendAttributes() allows to add attributes to the 
# outputed UI element
plotOutput(
  "plot"
) %>% tagAppendAttributes(
  onclick = "alert('hello world')"
)
```

Here is, for example, a small `{shiny}` app that implements this behavior:

```{r, Shiny with Magrittr, eval = FALSE}
library(shiny)
library(magrittr)
ui <- function(){
  fluidPage(
    # We create a plotOutput, which will show an alert when 
    # it is clicked
    plotOutput(
      "plot"
    ) %>% tagAppendAttributes(
      onclick = "alert('iris plot!')"
    )
  )
}
server <- function(input, output, session){
  output$plot <- renderPlot({
    plot(iris)
  })
}
shinyApp(ui, server)
```

**Lets put this into Golem!!!**

Use the `add_js_file("name")` function in Golem.

- In `/inst/app/www/script.js`

```{js, Golem Javascript example, eval=FALSE}
function alertme(id){
  // Asking information
  var name = prompt("Who are you?");
  // Showing an alert
  alert("Hello " + name + "! You're seeing " + id);
}
```

- Then in R

```{r, Golem R Code Example, eval=FALSE, echo=TRUE}
plotOutput(
  "plot"
) %>% tagAppendAttributes(
  # Calling the function which has been defined in the 
  # external script
  onclick = "alertme('plot')"
)
```

Last example, you can also add jQuery calls too! Inside this `inst/app/www/script.js`, you can also attach a new behavior with `jQuery` to one or several elements.
For example, you can add this `alertme` / `onclick` behavior to all plots of the app:

```{js, jQuery example, eval = FALSE, echo = TRUE}
function alertme(id){
  var name = prompt("Who are you?");
  alert("Hello " + name + "! You're seeing " + id);
}
/* We're adding this so that the function is launched only
when the document is ready */
$(function(){ 
  // Selecting all `{shiny}` plots
  $(".shiny-plot-output").on("click", function(){
    /* Calling the alertme function with the id 
    of the clicked plot */
    alertme(this.id);
  });
});
```

We've unpacked a lot here...but there is more. Check out [JavaScript Events in `{shiny}`](https://shiny.rstudio.com/articles/js-events.html) for the full list of JavaScript events available in `{shiny}`.

## JavaScript <-> `{shiny}` Communication
Now that we have seen some client-side optimization, i.e. R does not do anything with these events when they happen (in fact R is not even aware they happened).

### From R to JavaScript
Calling JS from the server side (i.e. from R) is done by defining a series of `CustomMessageHandler` functions: these are functions with one argument that can then be called using the `session$sendCustomMessage()` method from the server side.Or if you are using `{golem}`, using the `invoke_js()` function.

Using the skeleton from `golem::add_js_handler("first_handler")`, we get this snippet:

```{r, Another jQuery snippet, eval=FALSE}
$( document ).ready(function() {
  Shiny.addCustomMessageHandler('fun', function(arg) {
  
  })
});
```

Then, you can call it from the server side:

```{r, Server Side Script Example, eval = FALSE}
session$sendCustomMessage("fun", list())
# OR
golem::invoke_js("fun", ...)
```

>NOTE: the `list()` argument will be coverted to JavaScript Object Notation (JSON) and read via JavaScript. If we have function `x` and variables `a` and `b`, then we can use the dot notation of `x.a` and `x.b` in our JavaScript.

For Example:

```{js, Dot Notation, eval = FALSE}
// We define a handler called "computed", that can be called 
// from the server side of the {shiny} application
Shiny.addCustomMessageHandler('computed', function(mess) {
  // The received value (in mess) is serialized in JSON, 
  // so we can  access the list element with object.name
  alert("Computed " + mess.what + " in " + mess.sec + " secs");
})
```

And within R:

```{r, Server side using Dot Notation, eval = FALSE}
observe({
  # Register the starting time
  deb <- Sys.time()
  # Mimic a long computation
  Sys.sleep(
    sample(1:5, 1)
  )
  # Calling the computed handler
  golem::invoke_js(
    "computed", 
    # We send a list, that will be turned into JSON
    list(
      what = "time", 
      sec = round(Sys.time() - deb)
    )
  )
})
```

### From JavaScript to R

How can you do the opposite (send JavaScript calls to R)?

There is an object called `Shiny` within your browser. This object can be used to send values to R by creating an `InputValue`.

```{js, JavaScript to R, eval = FALSE}
// This function from the Shiny JavaScript object
// Allows to register an input name, and a value
Shiny.setInputValue("rand", Math.random())
```

You bind an input which can be caut fro the server side using:

```{r, JavaScipt to R Example 2, eval = FALSE}
# Once the input is set, it can be caught with R using:
observeEvent( input$rand , {
  print( input$rand )
})
```

`Shiny.setInputValue` cn be used inside any JavaScript function. As an example we use the following code snippet:

- In `inst/app/www/script.js`

```{r, JavaScript Example 3, eval = FALSE}
function alertme(){
  var name = prompt("Who are you?");
  alert("Hello " + name + "! Welcome to my app");
  Shiny.setInputValue("username", name)
}

$(function(){ 
  // Waiting for `{shiny}` to be connected
  $(document).on('shiny:connected', function(event) {
    alertme();
  });
  
  $(".shiny-plot-output").on("click", function(){
    /* Calling the alertme function with the id 
    of the clicked plot. 
    The `this` object here refers to the clicked element*/
    Shiny.setInputValue("last_plot_clicked", this.id);
  });
});

```

The above snippet gets the user name and last plot clicked. We can catch the call to the server using the following snippet:

```{r, JavaScript Receipt to Server, eval = FALSE}
# We wait for the output of alertme(), which will set the
# "username" input value
observeEvent( input$username , {
  cli::cat_rule("User name:")
  print(input$username)
})

# This will print the id of the last clicked plot
observeEvent( input$last_plot_clicked , {
  cli::cat_rule("Last plot clicked:")
  print(input$last_plot_clicked)
})
```

Which will give:

```{r, Server Output, eval = FALSE}
> golex::run_app()
Loading required package: shiny

Listening on http://127.0.0.1:5495
── User name: ─────────────────────────────────────────────────────
[1] "Colin"
── Last plot clicked: ─────────────────────────────────────────────
[1] "plota"
── Last plot clicked: ─────────────────────────────────────────────
[1] "plotb"
```


> NOTE: If you are using modules, you will need to pass the namespacing of the `id` to be able to get it back from the server. This can be accomplished using the `session$ns` function. This comes by default using any golem-generated module.

Example of a `session$ns` call used in Golem Module:

```{js, Golem Session Namespace use, eval = FALSE}
$( document ).ready(function() {
  // Setting a custom handler that will
  // ask the users their name
  // then set the returned value to a Shiny input
  Shiny.addCustomMessageHandler('whoareyou', function(arg) {
    var name = prompt("Who are you?")
    Shiny.setInputValue(arg.id, name);
  })
});
```

```{r, Golem Session Namespace use 2, eval = FALSE}
mod_my_first_module_ui <- function(id){
  ns <- NS(id)
  tagList(
    actionButton(
      ns("showname"), "Enter your name"
    )
  )
}

mod_my_first_module_server <- function(input, output, session){
  ns <- session$ns
  # Whenever the button is clicked, 
  # we call the CustomMessageHandler
  observeEvent( input$showname , {
    # Calling the "whoareyou" handler
    golem::invoke_js(
      "whoareyou", 
      # The id is namespaced, 
      # so that we get it back on the server-side
      list(
        id = ns("name")
      )
    )
  })
  
  # Waiting for input$name to  be set with JavaScript
  observeEvent( input$name , {
    cli::cat_rule("Username is:")
    print(input$name)
  })
}
```

## About {shinyjs} JS functions
As mentioned at the beginning of the chapter, running JavaScript code you don't fully control/understand can be tricky and may open vulnerabilities. It is always safer to use proven packages that have been tested and passed efficiency. For example [`shinyjs`](https://deanattali.com/shinyjs/).

See [deanattali.com/shinyjs/](https://deanattali.com/shinyjs/) for more information.

## One last thing: API calls
If you ShinyApp uses API calls, there are good chances you are already doing this from R. However, there may be downsides to this approach. As an example, if you are limited by API calls (Twitter API, Google Map API). Due to these restrictions your end user may be limited in using your app.

A work around is instead of using your Server's IP to make the API call, wrap it in Javascript so you use your user's IP instead! This allows your ShinyApp to scale.

You use the `fetch()` JavaScript function to make API calls. For an example: `tags$button("Get Me One!", onclick = "get_rand_beer()")`.

Below is a general skeleton that works with an API that does not require authentication and returns a JSON object.

- Inside JavaScript (here, we create a function that will be available on an onclick event):

```{js, Get a Beer, eval = FALSE}
// FUNCTION definition
const get_rand_beer = () => {
  // Fetching the data
  fetch("https://api.punkapi.com/v2/beers/random")
  // What do we do when we receive the data
  .then((data) =>{
  // TTurn the data to JSON
    data.json().then((res) => {
    // Send the json to R
      Shiny.setInputValue("beer", res, {priority: 'event'})
    })
  })
  // Define what happens if we fail to fetch
  .catch((error) => {
      alert("Error catching result from API")
  })
};
```

- Observe the event in your server:

```{r}
observeEvent( input$beer , {
  # Do things with beer
})
```

>NOTE: The data shared between R and JavaScript is serialized to JSON, so you will have to manipulate the format once you recieve it in R.

Learn more about `fetch()` at [Using Fetch](https://developer.mozilla.org/en-US/docs/Web/API/Fetch_API/Using_Fetch).

## Learn more about JavaScript

If you want to interact straight with R using NodeJS, you can use the `{bubble}` package. Obviously, you need to have NodeJS installed on your machine to use this service.

Install `{bubble}` from GitHub:

```{r, Install Bubble, eval = FALSE}
remotes::install_github("ColinFay/bubble")
```

You can use it in RMarkdown chunks by setting the `knitr` engine:

```{r, Configure Knitr, eval = FALSE}
bubble::set_node_engine()
```

Or straight from command line with:

```{r, Node Commandline, eval = FALSE}
node_repl()
```

Here is a list of external resources to learn more about JavaScript:

### `{shiny}` and JavaScript

-   We have written an online, freely available book about `{shiny}` and JavaScript: [_JavaScript 4 `{shiny}` - Field Notes_](http://connect.thinkr.fr/js4shinyfieldnotes/).

-   [JavaScript for `{shiny}` Users](https://js4shiny.com/), companion website to the rstudio::conf(2020) workshop.

-   [Build custom input objects](https://shiny.rstudio.com/articles/building-inputs.html).

-   [Packaging JavaScript code for `{shiny}`](https://shiny.rstudio.com/articles/packaging-javascript.html).

-   [Communicating with `{shiny}` via JavaScript](https://shiny.rstudio.com/articles/communicating-with-js.html).

### JavaScript basics

-   [Mozilla JavaScript](https://developer.mozilla.org/en-US/docs/Web/JavaScript)
-   [w3schools JavaScript](https://www.w3schools.com/js/default.asp)
-   [Free Code Camp](https://www.freecodecamp.org/)
-   [JavaScript for Cats](http://jsforcats.com/)
-   [Learn JS](https://www.learn-js.org/)

### jQuery

-   [jQuery Learning Center](https://learn.jquery.com/)
-   [w3schools jQuery](https://www.w3schools.com/jquery/default.asp)

### Intermediate/advanced JavaScript

-   [Eloquent JavaScript](https://eloquentjavascript.net/)
-   [You Don't Know JS Yet](https://github.com/getify/You-Dont-Know-JS)


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/XHz5s_T6MQU")`

<details>
  <summary> Meeting chat log </summary>
```
00:10:56	Ryan Metcalf:	This is a side project I’m working through…..nothing of presentation yet: https://xapi.com/overview/?utm_source=google&utm_medium=natural_search
00:19:45	Federica Gazzelloni:	@Russ can you put the title of the book?
00:19:59	russ:	"Javascript for R"
00:20:08	russ:	John Coene
00:20:09	Federica Gazzelloni:	is there any pdf or free available resource for that?
00:21:04	russ:	https://book.javascript-for-r.com/
00:21:23	Federica Gazzelloni:	thanks!
00:48:26	russ:	Standard generalized markup language
01:13:24	Federica Gazzelloni:	https://shiny.rstudio.com/articles/communicating-with-js.html
01:18:02	Federica Gazzelloni:	little example app that continuously sends rainbow colors from R to JavaScript
```
</details>
