# A Gentle Introduction to CSS

**Learning Objectives**

 - What CSS does
 - CSS is already used by {shiny}
 - How to write custom CSS
 - Integration of custom CSS with {shiny}

## What is CSS

CSS

- "Cascading style sheets"
- Encodes webpage design
  - View a page without CSS ("Web Developer" -> CSS -> Disable All Styles)

### Default CSS for {shiny}

Without fluidPage:

```{r css-1}
library(shiny)

my_tags <- tagList(
  h1("Hey"), 
  h2("You"), 
  p("You rock!"), 
  selectInput("what", "Do you", unique(iris$Species))
)

ui1 <- function(){
  my_tags
}

ui2 <- function(){
  fluidPage(my_tags)
}

empty_server <- function(input, output, session){}
```

```r
shinyApp(ui1, empty_server)

shinyApp(ui2, empty_server)
```

Show the HTML

```{r css-2}
print(ui1())

print(ui2())
```

## Getting started with CSS

- HTML is packed full of tags
- You use CSS to explain how each HTML element should look
- If the element has this tag ... it should look like this

```css
h2 {
  color: red;
}

//
selector {
  property: value;
}
```

Selecting elements based on different types of properties:

```css
// examples from ui2() above

// eg `h2` or `div`
the-name {
  property: value;
}

// eg, `#what`
#the-id {
  property: value;
}

// eg, `.control-label`
.the-class {
  property: value;
}
```

You can combine the selectors together:

```
h1,h2    // either `h1` or `h2`
div h1   // an `h1` inside a `div`
div > h1 // an `h1` immediately inside a `div`
a:hover  // an `a`, but only when the user hovers over it
```

and so on.

See the wonderful tutorial at MDN for more on the various ways to select elements:

[https://developer.mozilla.org/en-US/docs/Learn/CSS/Building_blocks/Selectors]()

## Integrate CSS files in your {shiny} app

CSS can be included in shiny apps in a few ways:

- inline
- in your `ui` definition
- in an imported file

### Inline

```{r css-3}
ui3 <- function(){
  tagList(
    h2(style = "color:red;", "This is red")
  )
}
```

```r
shinyApp(ui3, empty_server)
```

```{r css-4}
print(ui3())
```

### Using tags$style()

```{r css-5}
ui4 <- function(){
  tagList(
    tags$style(
      "h2{
        color:red;
      }"
    ), 
    h2("This is red")
  )
}
```

```{r css-6}
print(ui4())
```

### Using external files

Provided you have this in the path `./www/style.css`

```css
h2 {
    color: red;
}
```

Then the following will incorporate that css into your vanilla shiny app.

```{r css-7}
ui5 <- function() {
  tagList(
    tags$link(
      rel = "stylesheet",
      type = "text/css",
      href = "www/style.css"
    ), 
    h2("This is red")
  )
}
```

A golem-based shiny app has the command `golem_add_external_resources()` in the UI definition. This
function pulls in any CSS or JS files. But, to add those CSS / JS files you have to use the correct
golem workflow:

```r
# For CSS
golem::add_css_file("style")
```

This adds the file `inst/app/www/style.css` to your package.

There are alternative ways to include CSS. In "Outstanding User Interfaces with Shiny", the author
recommends creating a dependency, and then including it as a tag in your UI:

```r
# R/style.R
css_dep <- function() {
  htmlDependency(
    name = "css dep", version = "1.0", src = path_to_css, stylesheet = filename
  )
}

# In ui.R
... tagList(tag, css_dep)
```

## External Resources

Tools

- "Web Developer" extension
  [https://chrome.google.com/webstore/detail/web-developer/bfbameneiokkgbdmiekhjnmfkcnldhhm/related]()

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/i7utiiM8fr8")`

<details>
  <summary> Meeting chat log </summary>
```
LOG
```
</details>
